Inferring Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-07-27 doc/paper-lcr.tex
added figures

Disturbingly, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}.Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_{\rm NFW} (r)}{1+\left(r/r_{t}\right)^2},
\ee
which is the same as the NFW profile in the limit that $r_t$ goes to infinity. Defining $x$ as the dimensionless projected radial distance $R/r_{s}$ and $\tau$ as the dimensionless truncation radius $r_{t}/r_{s}$, \citet{BMO} derives the projected mass density, which is given by:
\begin{align}\label{eq:kappaBMO}
\Sigma_{\rm BMO}(x) = {2\tau^2 \over (\tau^2+1)^2}\left(
        {\tau^2+1\over x^2-1}\left(1-\mathcal{F}(x)\right)
        +
        2\mathcal{F}(x)\right.\nonumber\\
        -
        \left. {\pi \over \sqrt{\tau^2+x^2}}
        +
        {(\tau^2-1)\mathcal{L}(x,\tau)
        \over
        \tau\sqrt{\tau^2+x^2}}\right)
\end{align}
where $\mathcal{F}(x)$ and $\mathcal{L}(x,\tau)$ are defined as
\be\label{eq:F} 
\mathcal{F}(x) \equiv \begin{cases}  \frac{{\rm cos}^{-1} (1/x)}{\sqrt{x^2-1}} \hspace{0.2cm} & (x>1) \\
\\
\frac{4-x^2}{3}  & (x=1)\\
\\
\frac{ {\rm cosh}^{-1} (1/x)}{\sqrt{1-x^2}} & (x<1)
\end{cases}
\ee
\be\label{eq:L}
\mathcal{L}(x,\tau) = \ln\left(\frac{x}{\sqrt{\tau^2+x^2}+\tau}\right)
\ee
%the convergence is given by $\Sigma_{\rm BMO}(x) / \Sigma_{\rm cr}(z)$.

@@

\tom{this paragraph is phrased badly}
If $\kapparec$ is to be used in a real survey it is important that it is not a biased estimator of the true $\kappax$. 
Ideally one wants $\langle \kapparec-\kappatrue \rangle$ to equal 0 and $\kapparec-\kappatrue$ to be independant of
$\kapparec$., since the void correction ensures $\langle \kapparec \rangle = 0$ any bias in $\langle \kapparec-\kapparay \rangle$ comes
from the variations in $\kapparay$. With 8.2$\times 10^8$ pixels, uniformly spaced over 196 deg$^2$ of simulated sky,
 $\langle \kapparay \rangle$ = -0.00014. However, it is improbable that such a large number of lines of sight
 could be reconstructed in reality and the mean of $\kappatrue$ for a small subset of sightlines will not necessarily
equal zero. We find this error to be
\be\label{eq:sigmav}
\sigma(\kappav) \simeq 0.036 / N^{\frac{1}{2}},
\ee
 where $N$ is the number of lines of sight, although $\pr(\kappav)$ has positive skew for small $N$.
 Care must be taken that the callibration lines of sight are not biased in $\kappax$ or the final $\pr(\kappav)$ will also be biased. For the rest of this work we use 1000  randomly chosen calibration lines of sight to calculate $\kappav$, introducing an error on $\kappax$ of $0.001$. 1000 callibration lines of sight is unfeasable currently, but should be reasonable in the LSST era. Since the true $\langle \kapparay \rangle$ = -0.00014 is two orders of magnitude smaller than intrinsic error of the halo prescription, $\kapparec$ can be seen as unbiased given a large enough sample of lines of sight. For an individual line of sight $\kapparec$ can also be shown
\begin{figure}\label{fig:magcut}
\includegraphics[width=\columnwidth]{mag_scatter.eps}
\end{figure}
\begin{figure}\label{fig:radcut}
\includegraphics[width=\columnwidth]{radius_scatter.eps}
\end{figure}

@@

>>>>

Disturbingly, if one integrates the density profile of an NFW profile, the total mass is divergent.
Similarly, if the universe is homogeneously populated with NFW halos, the projected surface mass along any
line sight will also be divergant\footnote{At large radius, $\Sigma_{\rm nfw} \propto R^{-2}$, but the differential number of halos centred within an annulus of width $\dee R$ is given by $\dee N_{\rm annulus} \propto R \dee R$, so  $\Sigma_{\rm total} \propto \int_{0}^{\infty} R^{-1} \dee R$, which diverges logarithmically}.Since infinite mass is unphysical, the profile must be truncated at some point. \flag{previous truncation results}. Several truncation profiles have been suggested \citep[e.g][]{BMO}, but beyond several virial radii, the amount of matter associated with a halo is likely to be low; \citet{Nandra2012} showed that the Universe's expansion can place an upper limit on the radius of NFW halos -- for cluster scale halos, this happens at around 25 times the NFW scale radius. In \Sref{sec:knownMh+z} we investigate how different truncations affect the recovered total convergence for a line of sight, using the truncated NFW profile of \citet{BMO}
\be\label{eq:bmoprofile}
\rho(r) = 
\frac{\rho_{\rm NFW} (r)}{1+\left(r/r_{t}\right)^2},
\ee
which is the same as the NFW profile in the limit that $r_t$ goes to infinity. Defining $x$ as the dimensionless projected radial distance $R/r_{s}$ and $\tau$ as the dimensionless truncation radius $r_{t}/r_{s}$, \citet{BMO} derives the projected mass density, which is given by:
\begin{align}\label{eq:kappaBMO}
\Sigma_{\rm}(x) = {2\tau^2 \over (\tau^2+1)^2}\left(
        {\tau^2+1\over x^2-1}\left(1-\mathcal{F}(x)\right)
        +
        2\mathcal{F}(x)\right.\nonumber\\
        -
        \left. {\pi \over \sqrt{\tau^2+x^2}}
        +
        {(\tau^2-1)\mathcal{L}(x,\tau)
        \over
        \tau\sqrt{\tau^2+x^2}}\right)
\end{align}
where $\mathcal{F}(x)$ and $\mathcal{L}(x,\tau)$ are defined as
\be\label{eq:F} 
\mathcal{F}(x) \equiv \begin{cases}  \frac{{\rm cos}^{-1} (1/x)}{\sqrt{x^2-1}} \hspace{0.2cm} & (x>1) \\
\\
\frac{4-x^2}{3}  & (x=1)\\
\\
\frac{ {\rm cosh}^{-1} (1/x)}{\sqrt{1-x^2}} & (x<1)
\end{cases}
\ee
\be\label{eq:L}
\mathcal{L}(x,\tau) = \ln\left(\frac{x}{\sqrt{\tau^2+x^2}+\tau}\right)
\ee
%the convergence is given by $\Sigma_{\rm BMO}(x) / \Sigma_{\rm cr}(z)$.

@@

\begin{figure}\label{fig:ScattervsTruncation}
\includegraphics[width=\columnwidth]{truncation_scatter.eps}
\caption[magcut]{}
\end{figure}

\tom{this paragraph is phrased badly}
If $\kapparec$ is to be used in a real survey it is important that it is not a biased estimator of the true $\kappax$. 
Ideally one wants $\langle \kapparec-\kappatrue \rangle$ to equal 0 and $\kapparec-\kappatrue$ to be independant of
$\kapparec$., since the void correction ensures $\langle \kapparec \rangle = 0$ any bias in $\langle \kapparec-\kapparay \rangle$ comes
from the variations in $\kapparay$. With 8.2$\times 10^8$ pixels, uniformly spaced over 196 deg$^2$ of simulated sky,
 $\langle \kapparay \rangle$ = -0.00014. However, it is improbable that such a large number of lines of sight
 could be reconstructed in reality and the mean of $\kappatrue$ for a small subset of sightlines will not necessarily
equal zero. We find this error to be
\be\label{eq:sigmav}
\sigma(\kappav) \simeq 0.036 / N^{\frac{1}{2}},
\ee
 where $N$ is the number of lines of sight, although $\pr(\kappav)$ has positive skew for small $N$.
 Care must be taken that the callibration lines of sight are not biased in $\kappax$ or the final $\pr(\kappav)$ will also be biased. For the rest of this work we use 1000  randomly chosen calibration lines of sight to calculate $\kappav$, introducing an error on $\kappax$ of $0.001$. 1000 callibration lines of sight is unfeasable currently, but should be reasonable in the LSST era. Since the true $\langle \kapparay \rangle$ = -0.00014 is two orders of magnitude smaller than intrinsic error of the halo prescription, $\kapparec$ can be seen as unbiased given a large enough sample of lines of sight and Figure \ref{fig:bias} shows that there is no statistically significant evidence that $\kapparec$ is a biased estimator, regardless of the individual value of $\kapparec$.

\begin{figure}\label{fig:isitbiased}
\includegraphics[width=\columnwidth]{isitbiased.eps}
\caption[Biased?]{}
\end{figure}
\begin{figure}\label{fig:magcut}
\includegraphics[width=\columnwidth]{mag_scatter.eps}
\caption[magcut]{}
\end{figure}
\begin{figure}\label{fig:radcut}
\includegraphics[width=\columnwidth]{radius_scatter.eps}
\caption[radius cut]{}
\end{figure}

@@

Real attempts to reconstruct the convergence along a line of sight will not have direct access to the masses of dark matter halo, 
and will likely not have access to their spectroscopic redshift either. Instead these properties must be inferred from astrophysical
observables. In principle the only observables are astrometric positions, photometric colours and possibly spectra. In this section we
quantify the errors induced by inferring the halo mass and redshift from the observables. Much work has already focused on using
photometric colours to infer stellar mass \citep[\eg][]{augerstellarpops} and redshifts \citep[\eg][]{BPZ}, in line with these we shall use
stellar mass and photometric redshifts (with appropriate errors), as the observables that must be converted into an estimate
of $\kappax$. We will investigate three main sources of error: inferring halo mass given observed stellar mass; halos at the wrong redshift
due to photometric redshift error, and; the combination of both redshift and stellar mass errors.
